Computing MEM seeding probabilities

### 2019-02-15 MEM.tex
Optimizing the figures and the text.

\begin{abstract}
High throughput sequencing produces massive amounts of data that must be
processed efficiently. Most reads are identified by mapping them to a
reference, typically a genome. Modern mappers rely on heuristics to gain
speed at a reasonable cost of accuracy. MEM seeding is a popular heuristic
that shows good empirical performance but for which no theory is
available. It is thus difficult to calibrate or optimize mapping
strategies based on MEM seeding. Here we develop a theory to predict the
performance of MEM seeding. More specifically, we design a method to
compute the probability a read is mapped to the right location when using
MEM seeds. Our computational strategy is based on the analytic
combinatorics programme, but uses only part of it. By recoding the reads
in a specialized combinatorial alphabet, we can implicitly specify their
generating function and extract the probabilities of interest by direct
methods or by Monte Carlo sampling. This yields an efficient algorithm to
calibrate the MEM seeding heuristic for short read technologies.
\end{abstract}

@@

Suppose that a DNA fragment drawn at random from a known
genome\footnote{We will assume throughout that the sequence of the genome
is known exactly and with absolute certainty. This is not the case in
practical application, but this assumption will allow us to develop a
theory centered on the read.} is sequenced with an imperfect instrument;
how can we identify the location of the fragment in the genome?

We will refer to this question as the \emph{true location problem}.
It is central to countless applications, including discovering
disease-causing mutations; diagnosing viral, bacterial or fungal
infections; diagnosing cancer and choosing appropriate therapies;
selecting breeds of interest in agriculture; tracing human migrations from
ancient bones; identifying the donor of a biological sample; diagnosing
genetic diseaes at birth; finding compatible graft donors \textit{etc}.

The focus of this document is not to solve the true location problem
\textit{per se}, but rather to evaluate whether a given solution is
correct. At first sight, whether we can map a read to its true location
seems to depend only on its length and on the error rate of the
instrument. However, there is another important factor to consider: most
genomes contain repeated sequences, \textit{i.e.} relatively large
subsequences that are present at more than one location. So we cannot map
the fragment with certainty if there exists an exact copy of the sequence
somewhere else in the genome, because we cannot know which of the two
copies was sequenced.

This case is rare though. Repeated sequences are in general not identical,
but merely homologous\footnote{By that we mean that drawing nucleotides
uniformly at random is very unlikely to produce sequences with the
observed degree of identity.}. So when the DNA fragment is duplicated, we
can map it with high confidence provided we can distinguish it from its
duplicates. This in turn depends on their similarity with the target
fragment and on the error rate of the sequencer.

@@

We will refer to this as the \emph{best location problem}. It amounts to
finding the optimal alignment between two sequences, and for this reason
has received substantial attention in bioinformatics. The distinction is
relevant for the present discussion because the candidate solution of the
\emph{true} location problem can only be the \emph{best} location. So in
order to solve the true location problem, we must first solve the best
location problem.

@@

There exists exact algorithms to solve the best location
problem\cite{pmid7265238,pmid5420325}, but they are too slow to process
the large amount of data generated by modern sequencers. Instead, one uses
heuristic methods, \textit{i.e.} algorithms that run fast, but may not
return the best location\cite{Waterman1984}. Heuristic mapping algorithms
thus have three possible outcomes: $i.$ the DNA fragment is mapped to the
best location, $ii.$ the DNA fragment is mapped to another location,
$iii.$ the DNA fragment is not mapped at all.

@@

If seeds can be discovered fast, we can quickly zoom in on a small set of
candidate locations and reduce the input size of the exact but slow
algorithm. The disadvantage is that the target location is not always in
the candidate set. In this case, it cannot be discovered and the outcome
is either a false positive or a false negative.

This imposes a trade-off between speed and accuracy between each candidate
location must be tested with a slow algorithm. If the seeding step is set
to produce many candidates, the target is likely to be among them and it
will be discovered. The downside is that this will take long. Conversely,
if the seeding step is set to produce few candidates, the process will
run faster but the target is less likely to be in the candidate list so it
is less likely to be discovered.

Importantly, the terms of this trade-off are not set in stone. Some
seeding methods are better than others\cite{pmid16533404,pmid20460430}, so
one can develop faster mapping algorithms at no cost for accuracy.

This framework was developed for the best location problem, but it also
applies to the true location problem. Actually, the process is exactly the
same but the outcomes $i.$ and $ii.$ are different. The first, called a
``true positive'' or an ``on-target hit'', is that the DNA fragment is
mapped to the \emph{true} location; the second, called a ``false
positive'' or an ``off-target hit'' is that the DNA fragment is mapped
somewhere else in the genome. A true positive thus requires that the
target (the true location) is in the candidate set, and that it has the
best score. For now, we will assume that the true location is always the
best and we defer the discussion about this assumption until
section~\ref{sec:not_the_best}.

@@

The rest of the document will focus on a seeding heuristic known as ``MEM
seeding''. This approach gives good empirical results when mapping short
reads (it is for instance the strategy\footnote{To be exact, BWA-MEM uses
SMEM seeds (Super Maximal Exact Match seeds).} of the popular mapper
BWA-MEM\cite{li2013aligning}) but there is presently no theory to justify
this performance. A deeper understanding of MEM seeding would thus be
useful to calibrate the speed/accuracy trade-off of this heurisitic.

Before developing a formal theory, let us clarify the meaning of the MEM
acronym.

@@

This definition presents a computational challenge: to know if a given
subsequence is a MEM seed, we need to know if it exists somewhere in the
genome. This is a non trivial problem in itself, but fortunately there
exists practical methods to identify all the MEM seeds from a read, even
for very large genomes\cite{ferragina2000opportunistic,
ferragina2005indexing}. Such algorithms are crucial for the present
discussion, but describing them falls outside the scope of the document.
We refer the interested reader to the relevant
literature\cite{pmid24336412,pmid25399029,pmid23349213,pmid19389736,
li2013aligning}, and here on we will assume that MEM seeds are simply
available at all times. Looking ahead to section~\ref{sec:est}, let
us just say that this can be done with an algorithm called the
\emph{backward search}\cite{ferragina2000opportunistic}.

Before going further, we need to move a practical consideration out of the
way. It stands to reason that very short matches between the read and the
genome are uninformative. For instance, $99.7\%$ of all the possible
12-mers are present in the reference human genome, so an exact match of
size 12 or lower says nothing about the location of the DNA fragment.
In what follows, we will assume that subsequences of the read can only
match the original DNA sequence and/or its duplicates, but not the rest of
the genome. This will simplify the discussion and we defer the discussion
of this assumption until section~\ref{sec:short_MEMs}

Fig.~\ref{fig:MEM_example} shows a concrete example. The target (the true
location of the read) is first potential location. The locations below
correspond to duplicate sequences. The mismatches between the genome and
the read are highlighted in bold. The read has three errors indicated by
stars, so \texttt{T}, \texttt{T} and \texttt{C} should actually be
\texttt{A}, \texttt{A} and \texttt{G}.

\begin{figure}[h]
\centering
\includegraphics[scale=1]{MEM_example.pdf}
\caption{\textbf{A concrete example of MEM seeding.}
A DNA fragment is sequenced with an imperfect instrument. It comes from
the top genomic location (Truth), but three errors were introduced
(indicated by stars). The sequence of the DNA fragment has three
duplicates at other genomic locations, shown below the target. Mismatches
between the read and the candidate sequence are highlighted in bold. MEM
seeds are indicated by grey boxes.}
\label{fig:MEM_example}
\end{figure}

@@

\begin{definition}
Assume that the thresohld $\gamma$ is fixed. A \emph{good} seed is an
on-target MEM seed of size $\gamma$ or greater. Every other MEM seed is a
\emph{bad} seed. In other words, the true location is in the candidate set
if and only if the read has a good seed.
\end{definition}

Compared to seeds of fixed size, MEM seeds have at least two
counter-intuitive properties. The first is that there are cases where
there is no good seed, even when $\gamma$ is decreased.
Fig.~\ref{fig:full_masking_example} shows such an example. Even though
there is a single sequencing error, the read does not have any MEM seed
matching the target. Lowering $\gamma$ does not change this, so there is
no way to discover the true location (even though it is also the best
location).

\begin{figure}[h]
\centering
\includegraphics[scale=1]{full_masking_example.pdf}
\caption{\textbf{MEM masking case 1.}
The read, the MEM seeds and the sequences are as in
Fig.~\ref{fig:MEM_example}. The MEM seeds matching the two incorrect
locations at the bottom effectively mask the target, so it cannot
be discovered. MEM masking can occur even when the true location is the
best and when there is a single error on the read.}
\label{fig:full_masking_example}
\end{figure}

The second counter-intuitive property of MEM seeding is that there are
cases where there would be a good seed if the read were shorter.
Fig.~\ref{fig:short_vs_long} shows an example of this case. Again, there
is no on-target MEM seed, but there would be if the read were two
nucleotides shorter on the right side. Indeed, in this case there would be
a shared MEM seed matching the first and the second locations. It would
also be a good seed for $\gamma \leq 12$.

\begin{figure}[h]
\centering
\includegraphics[scale=1]{short_vs_long_example.pdf}
\caption{\textbf{MEM masking case 2.}
The read, the MEM seeds and the sequences are as in
Fig.~\ref{fig:MEM_example}. There is no on-target MEM seed, but there
would be a shared MEM seed if the read were two nucleotides shorter. The
true location has been masked by sequencing the last two nucleotides.}
\label{fig:short_vs_long}
\end{figure}

These examples show that MEM seeds can perform worse than seeds of fixed
size, again raising the question why they seem to give good performance in
practice. The theory developed below will allow us to compute the
probability that a read has no good seed and thus that the true location
is missed. This probability is the key to evaluating and calibrating the
MEM seeding heuristic.

From now we will make the simplifying assumption that a read with a good
seed is always mapped at the correct location. In other words, we assume
that a good seed is necessary and sufficient for a true positive or
outcome $i.$ In practice this is not always the case, but this assumption
shifts the burden of mapping to the alignment algorithm and allows us
focus on improving the seeding heuristic.

We first explain a detailed strategy to compute the probablity of outcome
$i.$ We then explain how to compute the probability of outcome $iii.$,
\textit{i.e.} that the seeding step returns no hit. Combining these two
quantities allows us to compute the probability of interest, which is that
the read is mapped to an incorrect location or outcome $ii.$ Finally, we
provide a computationally efficient method for estimating the parameters
that are necessary to estimate the probabilities of interest.

Before going further, we recommend the readers that are not already
familiar with MEM seeding to take a moment to study the examples above.
The complexity of the next sections will go \textit{crescendo} and they
will draw heavily from the concepts and definitions above.

@@

For each potential location including the target, we define the
\emph{match streak} at a given nucleotide as the number of nucleotides
since the last mismatch on the left, or since the beginning of the read.
For instance, in Fig.~\ref{fig:MEM_example}, the match streak of the
target at the ninth nucleotide is $1$ while for the second potential
location, it is $7$.

\begin{definition}
At a given position of the read, an off-target location is a \emph{hard
mask} if its success streak is strictly longer than the match streak of
the target. An off-target location is a \emph{soft mask} if its has the
same match streak as the target.
\end{definition}

For instance, the second location in the example above is a hard mask at
the ninth nucleotide. On Fig.~\ref{fig:MEM_example}, one can further see
that at the second nucleotide, all the off-target locations are soft masks
and that at the last nucleotide, the second location is a soft mask.

From the definition, the last nucleotide of every strict on-target MEM
seed is always unmasked. Conversely, an unmasked nucleotide always belongs
to exactly one strict on-target MEM seed. Also, the last nucleotide of
every shared on-target MEM seed is always soft-masked, but a soft-masked
nucleotide does not always belong to a shared on-target MEM seed.

@@

\begin{figure}[h]
\centering
\includegraphics[scale=.85]{sketch_extended.pdf}
\caption{\textbf{The extended MEM encoding}.
The read of Fig.~\ref{fig:MEM_example} is represented in the MEM alphabet.
The arrows departing from those numbers help understand their meaning. The
$\downarrow$ symbol is decorated by the number of locations that match the
nucleotide. The $\uparrow$ symbol is decorated by the number of
nucleotides from the last error or from the beginning of the read. Also
note the presence of the terminator $|$ at the end of the read.}
\label{fig:sketch_extended}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[scale=0.88]{sketch_dual.pdf}
\caption{\textbf{The +/- encoding.}
A example read is represented in the +/- alphabet. The symbols
$\downarrow_{/j}^-$ represent a mismatch against the off-target sequence,
the symbols $\downarrow_{/j}^+$ represent a mismatch against the target,
and the symbol $\Downarrow$ represent a mismatch against both. The index
$j$ is the number of matches since the last mismatch for the other
sequence. Note the presence of the terminator $|$ at the end of the read.
}
\label{fig:dual}
\end{figure}

We use the same error model as before: for every nucleotide, the read
differs from the target with a fixed probability $p$ and the sequences
differ from each other with a fixed probability $\mu$. With these
definitions, the probability of occurrence of the symbol $\square$ is $a =
(1-p)(1-\mu)$. Likewise, the collective probability of occurrence of the
symbols $\downarrow_{/j}^-$ is $b = p\mu/3$, that of the symbols 
$\downarrow_{/j}^+$ is $c = (1-p)\mu$, and the probability of occurrence
of the symbol $\Downarrow$ is $d = p(1-\mu/3)$. It is easy to verify that
$a+b+c+d=1$.

@@

Assuming that there are $N = N_0$ duplicates and that we run $m$
iterations of the backward search, the probability of the sequence of
observations $(N_1, N_2, \ldots, N_m)$ is thus

@@

To avoid spending too much time on small random MEM seeds, we can
prioritize the seeds so that the best hit is likely to be discovered first
and bail out from the alignments as soon as it appears that their score
will be worse than the best hit so far. We can even put an upper bound on
the quality of candidate so that the alignment can be skipped altogether.
For instance, if we already have a candidate at distance 1 from the read,
there is no need to verify candidates from MEM seeds that do not reach one
end of the read. Indeed, the hit must be at distance at least 2 because
there is a mismatch before the MEM seed and another one after it.

@@

In general, one does not need to worry about the value of $\beta$ because
random hits are extremely visible when $p$ is small. For instance, one can
choose $\beta=1/2$ so that the term $(1-/\beta)/\beta$ disappears to
simplify expression (\ref{eq:bayes}). We compute this expression for every
read and add the result to the estimated probability of false positive for
the read. Either the term is negligible and it does not affect the
estimated probability, or it is strongly dominant and yields a very strong
probability that the location is incorrect, regardless of the estimate
from the MEM seeding process.

>>>>

\begin{abstract}
High throughput sequencing produces massive amounts of data that must be
processed efficiently. Most reads are identified by mapping them to a
reference, typically a genome. Modern mappers rely on heuristics to gain
speed at a reasonable cost of accuracy. MEM seeding (where MEM stands for
Maximal Exact Match) is a popular heuristic that shows good empirical
performance but for which no theory is available. It is thus difficult to
calibrate or optimize mapping strategies based on MEM seeding. Here we
develop a theory to predict the performance of MEM seeding. More
specifically, we design a method to compute the probability a read is
mapped to the right location when using MEM seeds. Our computational
strategy is based on the analytic combinatorics programme, but uses only
part of it. By recoding the reads in a specialized combinatorial alphabet,
we can implicitly specify their generating function and extract the
probabilities of interest by direct methods or by Monte Carlo sampling.
This yields an efficient algorithm to calibrate the MEM seeding heuristic
for short read technologies.
\end{abstract}

@@

Suppose that a DNA fragment drawn at random from a known
genome\footnote{We will assume that the sequence of the genome is known
exactly and with absolute certainty. This is not the case in practical
application, but this assumption will allow us to develop a theory
centered on the read.} is sequenced with an imperfect instrument; how can
we identify the location of the fragment in the genome?

We will refer to this question as the \emph{true location problem}.
It has applications in countless domains, like discovering
disease-causing mutations; diagnosing viral, bacterial or fungal
infections; diagnosing cancer and choosing appropriate therapies;
selecting breeds of interest in agriculture; tracing human migrations from
ancient bones; identifying the donor of a biological sample; diagnosing
genetic diseaes at birth; finding compatible graft donors \textit{etc}.

The focus of this document is not to solve the true location problem
\textit{per se}, but rather to evaluate whether a given solution is
correct. At first sight, whether we can map a read to its true location
seems to depend only on its length and on the error rate of the
sequencer. However, there is another important factor to consider: most
genomes contain repeated sequences, \textit{i.e.} relatively large
subsequences that are present at more than one location. For instance, we
cannot map the fragment with certainty if there exists an exact copy of
the sequence somewhere else in the genome, because we cannot know which of
the two copies was sequenced.

This case is rare though. Repeated sequences are in general not identical,
but merely homologous\footnote{By that we mean that drawing nucleotides
uniformly at random is very unlikely to produce sequences with the
observed degree of identity, meaning that the similarity is due some
biological duplication process.}. So when the DNA fragment is duplicated,
we can still map it, as long as we can distinguish it from its duplicates.
This in turn depends on their similarity with the target fragment and on
the error rate of the sequencer.

@@

We will refer to this task as the \emph{best location problem}. It amounts
to finding the optimal alignment between two sequences, and for this
reason has received substantial attention in bioinformatics. Finding the
best location of the read is the only way to get to its true location, so
we need to focus our attention on the best location problem.

@@

There exists exact algorithms to solve the best location
problem\cite{pmid7265238,pmid5420325}, but they are too slow to process
the large amount of data generated by modern sequencers. Instead, one uses
heuristic methods, \textit{i.e.} algorithms that run fast, but may not
return the best location\cite{Waterman1984}. Heuristic mapping algorithms
thus have three possible outcomes: $i.$ the DNA fragment is mapped to the
best location, $ii.$ the DNA fragment is mapped to some other location,
$iii.$ the DNA fragment is not mapped at all.

@@

If seeds can be discovered fast, we can quickly zoom in on a small set of
candidate locations and reduce the input size of the (slow) alignment
algorithm. The disadvantage is that the target location is not always in
the candidate set. In this case, it cannot be discovered and the outcome
is either $ii.$ or $iii.$

This imposes a trade-off between speed and accuracy because each candidate
location must be tested with a slow algorithm. If the seeding step is set
to produce many candidates, the target is likely to be among them and it
will be discovered. The downside is that this will take long. Conversely,
if the seeding step is set to produce few candidates, the process will
run faster but the target is less likely to be in the candidate list so it
is less likely to be discovered.

Importantly, the terms of this trade-off depend on the algorithm. Some
seeding methods are better than others\cite{pmid16533404,pmid20460430}, so
one can develop faster mapping algorithms at no cost for accuracy.

The seeding framework was developed for the best location problem, but it
also applies to the true location problem. Actually, the process is
exactly the same except that outcomes $i.$ and $ii.$ now refer to the true
location. Outcome $i.$, called a ``true positive'', is that the DNA
fragment is mapped to the \emph{true} location. Outcome $ii.$, called a
``false positive'' is that the DNA fragment is mapped to an incorrect
location. Outcome $iii.$, called a ``false negative'', does not change and
still means that the DNA fragment is not mapped at all.

We will make two assumptions that will simplify the discussion. The first
is that the true location is always the best. We will further comment on
this and explain how to relax this assumption in section
section~\ref{sec:not_the_best}. Until then, we consider that the true
location problem is the same as the best location problem. The second
assumption is that the best location is always discovered if it is in the
candidate set. This implies that outcomes $i.$, $ii.$ and $iii.$ are
determined by the seeding process. For this reason, we will sometimes
refer to them as the outcomes of the seeding step, even though they are
the outcomes of the whole mapping process.

@@

The rest of the document will focus on a seeding heuristic known as ``MEM
seeding'' (where MEM stands for Maximal Exact Match). This approach gives
good empirical results when mapping short reads (it is for instance the
strategy\footnote{To be exact, BWA-MEM uses SMEM seeds (Super Maximal
Exact Match seeds).} of the popular mapper BWA-MEM\cite{li2013aligning})
but there is presently no theory to justify its performance. A deeper
understanding of MEM seeding would be useful to calibrate the
speed/accuracy trade-off of this heurisitic.

Before developing a formal theory, let us explain in more detail what
Maximal Exact Matches are and how MEM seeding works.

@@

This definition presents a computational challenge: to know if a given
subsequence is a MEM seed we need to know if it exists somewhere in the
genome. This is a non trivial problem in itself, but fortunately there
exists practical methods to identify MEM seeds, even when the reference
genome is very large\cite{ferragina2000opportunistic,
ferragina2005indexing}. Such algorithms are crucial for the present
discussion, but describing them falls outside the scope of the document.
We refer the interested reader to the relevant
literature\cite{pmid24336412,pmid25399029,pmid23349213,pmid19389736,
li2013aligning}, and here on we will assume that MEM seeds are available
at all times. Looking ahead to section~\ref{sec:est}, let us just mention
that this can be done with an algorithm called the \emph{backward
search}\cite{ferragina2000opportunistic}.

Before going further, we need to move a practical consideration out of the
way. It stands to reason that very short matches between the read and the
genome are uninformative. For instance, $99.7\%$ of all the possible
12-mers are present in the reference human genome, so an exact match of
size 12 or lower says nothing about the location of the DNA fragment. In
what follows, we will assume that subsequences of the read can only match
the original DNA sequence and/or its duplicates, but not the rest of the
genome. This will greatly simplify the exposition of the theory. We will
discuss how to relax this assumption in section~\ref{sec:short_MEMs}.

Fig.~\ref{fig:MEM_example} shows a concrete example where a read is
aligned to candidate locations. The target (the true location of the
read) is first potential location. The locations below correspond to
duplicate sequences. The mismatches between the genome and the read are
highlighted in bold. The read has three errors indicated by stars, so
\texttt{T}, \texttt{T} and \texttt{C} should actually be \texttt{A},
\texttt{A} and \texttt{G} (\textit{i.e.} as in the target sequence).

\begin{figure}[h]
\centering
\includegraphics[scale=1]{MEM_example.pdf}
\caption{\textbf{Example of MEM seeding.}
A DNA fragment is sequenced with an imperfect instrument. It comes from
the top genomic location (Truth), but three errors were introduced
(indicated by stars). The sequence of the DNA fragment has three
duplicates at other genomic locations, shown below the target. Mismatches
between the read and the candidate sequence are highlighted in bold. MEM
seeds are indicated by grey boxes.}
\label{fig:MEM_example}
\end{figure}

@@

\begin{definition}
Assume that the thresohld $\gamma$ is fixed. A \emph{good} seed is an
on-target MEM seed of size $\gamma$ or greater. In other words, the true
location is in the candidate set if and only if the read contains a
good seed.
\end{definition}

Compared to seeds of fixed size, MEM seeds have at least two
counter-intuitive (and undesirable) properties. The first is that there
are cases where there cannot be any good seed, even when changing
$\gamma$. Fig.~\ref{fig:full_masking_example} shows such an example. Even
though there is a single sequencing error, the read has no good seed.
Lowering $\gamma$ does not change this, so there is no way to discover the
true location (even though it is the best location).

\begin{figure}[h]
\centering
\includegraphics[scale=1]{full_masking_example.pdf}
\caption{\textbf{Issues with MEM seeds: inaccessible targets.}
The read, the MEM seeds and the sequences are represented as in
Fig.~\ref{fig:MEM_example}. The MEM seeds matching the two incorrect
locations at the bottom effectively hide the target, so it cannot
be discovered. This can occur even when the true location is the
best and when there is a single error on the read.}
\label{fig:full_masking_example}
\end{figure}

The second counter-intuitive property of MEM seeding is that shortening
a read can sometimes create a good seed. Fig.~\ref{fig:short_vs_long}
shows an example of this case. Again, there is no good seed, but there
would be one if the read were two nucleotides shorter on the right side.
Indeed, in this case there would be a shared MEM seed matching the first
and the second locations, provided $\gamma \leq 12$.

\begin{figure}[h]
\centering
\includegraphics[scale=1]{short_vs_long_example.pdf}
\caption{\textbf{Issues with MEM seeds: too long reads.}
The read, the MEM seeds and the sequences are represented as in
Fig.~\ref{fig:MEM_example}. There would be a good seed (shared) if the
read were two nucleotides shorter. The true location is hidden by
the last two nucleotides.}
\label{fig:short_vs_long}
\end{figure}

These examples show that MEM seeds can perform worse than seeds of fixed
size. MEM seeds yield fewer candidates and therefore speed up the mapping
process, but the question is at what cost for accuracy? The theory
developed below will allow us to compute the probability that a read has
no good seed and thus that the true location is missed. This probability
is the key to evaluating and calibrating the MEM seeding heuristic.

We first expose a strategy to compute the probablity of true positives
(outcome $i.$), then show how to compute the probability of false
negatives (outcome $iii.$), and combine these two quantities to compute
the probability of false positives (outcome $ii.$). Finally, we provide a
computationally efficient method for estimating the parameters that are
necessary to estimate the probabilities of interest.

@@

For each potential location including the target, we define the
\emph{match streak} at a given nucleotide as the number of nucleotides
since the last mismatch on the left, or since the beginning of the read.

\begin{definition}
At a given position of the read, an off-target sequence is a \emph{hard
mask} if its success streak is strictly longer than the match streak of
the target. An off-target sequence is a \emph{soft mask} if its has the
same match streak as the target.
\end{definition}

Fig.~\ref{fig:hard_vs_soft_masks} illustrates hard and soft masks with an
example. It is important to bear in mind that hard and soft masks depend
on the position of interest: a sequence can be a mask at the left end of
the read but not at the right end, or \textit{vice versa}. In the example
of Fig.~\ref{fig:full_masking_example}, every nucleotide of the read has a
hard mask, which is why the target cannot be discovered. In the example of
Fig.~\ref{fig:short_vs_long}, the middle sequence is a soft mask on the
right half of the read, but it turns into a hard mask two nucleotides
before the right end, which is why the target could be discovered if the
read were shorter.

\begin{figure}[h]
\centering
\includegraphics[scale=0.85]{hard_vs_soft_masks.pdf}
\caption{\textbf{Example of hard and soft masks.}
The nucleotides that do not match the read are reprsented as black boxes.
The nucleotides that contribute to the match streak are represented as
grey boxes. The remaining nucleotides are represented as open boxes. At
the given position, the match streak of the target is 7. The first
off-target is a hard mask because its match streak is $13>7$. The second
off-target sequence is a soft mask because its match streak is 7 like the
target. The third off-target sequence is not a mask because its match
streak is $2<7$.}
\label{fig:hard_vs_soft_masks}
\end{figure}

From the definition, we see that the last nucleotide of every strict
on-target MEM seed is always unmasked. Conversely, an unmasked nucleotide
always belongs to exactly one strict on-target MEM seed. Also, the last
nucleotide of every shared on-target MEM seed is always soft-masked, but a
soft-masked nucleotide does not always belong to a shared on-target MEM
seed.

@@

\begin{figure}[h]
\centering
\includegraphics[scale=.85]{sketch_extended.pdf}
\caption{\textbf{The MEM encoding}.
The read of Fig.~\ref{fig:MEM_example} is represented in the MEM alphabet.
The arrows departing from those numbers help understand their meaning. The
$\downarrow$ symbol is decorated by the number of locations that match the
nucleotide. The $\uparrow$ symbol is decorated by the number of
nucleotides from the last error or from the beginning of the read. Also
note the presence of the terminator $|$ at the end of the read.}
\label{fig:sketch_extended}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[scale=0.85]{sketch_dual.pdf}
\caption{\textbf{The +/- encoding.}
A example read is represented in the +/- alphabet. The symbols
$\downarrow_{/j}^-$ represent a mismatch against the off-target sequence,
the symbols $\downarrow_{/j}^+$ represent a mismatch against the target,
and the symbol $\Downarrow$ represent a mismatch against both. The index
$j$ is the number of matches since the last mismatch for the other
sequence. Note the presence of the terminator $|$ at the end of the read.
}
\label{fig:dual}
\end{figure}

We use the same error model as before: for every nucleotide, the read
differs from the target with a fixed probability $p$ and the sequences
differ from each other with a fixed probability $\mu$. With these
definitions, the probability of occurrence of the symbol $\square$ is $a =
(1-p)(1-\mu)$. Likewise, the collective probability of occurrence of the
symbols $\downarrow_{/j}^-$ is $b = (1-p)\mu$, that of the symbols
$\downarrow_{/j}^+$ is $c = p\mu/3$, and the probability of occurrence of
the symbol $\Downarrow$ is $d = p(1-\mu/3)$. It is easy to verify that
$a+b+c+d=1$.

@@

Assuming that there are $N = N_0$ duplicates and that we run $m$
iterations of the backward search, the probability of the observations
$(N_1, N_2, \ldots, N_m)$ is thus

@@

To save time during the alignment phase, we can prioritize the seeds so
that the best hit is likely to be discovered first, allowing us to bail
out from other alignments as early as possible. We can even bound
the maximum quality of some candidates so that the alignment can be
skipped altogether. For instance, if we already have a candidate at
distance 1 from the read, there is no need to verify candidates from MEM
seeds in the middle of the read. Indeed, the hit must be at distance at
least 2 because there is a mismatch before the MEM seed and another one
after it.

@@

In general, one does not need to worry about the value of $\beta$ because
random hits are extremely visible when $p$ is small. For instance, one can
choose $\beta=1/2$ so that the term $(1-/\beta)/\beta$ disappears in
expression (\ref{eq:bayes}).

Because the probability that the best candidate is a random sequence is
either very small or very large, it either has no influence on the
probability of false positive, or completely dominates it. For every read,
we thus take the maximum of this probability or the estimated probability
that MEM seeding returned a false positive.
